---
title: "Module 4 Assignment"
author: "GEOG-3440"
output: html_document
---

```{r setup, include=FALSE} 
knitr::opts_chunk$set(echo = TRUE)
```

Be sure to **read assigned chapters and articles** and carefully walk through all of the **Module Guide** before starting on this task.

*Avoid using packages or solutions beyond the content covered in class or the guide.*

**Submit** your responses in a **single, organized html file written in RMardown.**

<br />


## **Module Tasks**

Complete the following tasks related to the content of this week. Code, text, and printed statements should be well-organized and clean. All plots should at contain useful axes labels and title. Avoid submitting messy and unnecessary information.


<br />

##### **Dates and time**

Perform the following datetime-related tasks. Keep in mind that datetime objects are recognized as distinct types of data.

1. Create a datetime object using the start *(Jan, 5)* and end *(April, 21)* dates of the spring 2026 semester. Use the `seq` and `as.Date` functions to create a daily series of all days in the semester, then print the `class` and `length` of the series.

2. Using the `weekdays` function, determine the percentage of days during the semester that are class days (i.e. Mondays and Wednesdays). Print the percentage *(no need to consider holidays or missed classes)*.

Run the following code to generate synthetic air temperature data:

```{r}
# generate 15 min time-series data for 4 days
start_time <- as.POSIXct("2026-01-28 00:00:00")
end_time   <- as.POSIXct("2026-01-30 23:45:00")
time <- seq(from = start_time, to = end_time, by = "15 min")

# create daily theoretical temperature variation with a range of 10 degrees F
daily_cycle <- 10 * sin(2 * pi * as.numeric(difftime(time, start_time, units = "hours")) / 24)

# vary temperature around 25 F and add noise (don't worry about understanding this at the moment)
value <- 25 + daily_cycle + rnorm(length(time), mean = 0, sd = 2.5)

# create data frame
df_temp <- data.frame( time = time, value = value)

# plot generate air temp
plot(df_temp$time, df_temp$value, type = "l", xlab = "Time", ylab = "Temperature (F)",
     main = "January Air Temperature at Field Site", col='orange')
```


3. Use the `format` function to create a new column in your data frame that converts all 15-min times to hourly datetime object (same length as your time variable). Then, use either the `aggregate` or `tapply` function to take the average air temperature each hour. Use this output to create a new plot that shows the 15 min air temperature in a dashed (parameter `lty`) light gray color, then add (use `lines()`) a solid red line to the same plot showing the average hourly values you computed. Include a legend indicating the two different data types.

*Read the help documents for each of the functions above and see the examples at the bottom to help of each to further help*

<br />



##### **Snow data**

Read in the csv file provided in canvas. This file contains daily snow and meteorological data for 30 days from December, 24, 2023 to January, 22 2024 collected at the snow telemetry (Snotel) station at the Timpanogos Divide.

For context, snow water equivalent is the amount of water in the snowpack or a volume of snow and is calculated by multiplying the snow density ($\rho$) as a percentage, by depth (*d*):

$$SWE = \rho * d$$
a brief explanation of other variables are also listed:

- date - Date string corresponding to local time (convert dates column to a date format)
- swe - Snow water equivalent at the start of the day (inches)
- swe_med - Historical median swe at the start of the day for 1991-2020 (inches)
- depth - Snow depth (inches)
- prec_accum - Total accumulated (cumulative) water for current water year (Oct 1st) (inches)
- prec_med - historical median prec_accum for 1991-2020 (inches)
- t_max - Air temperature maximum of the day ($^{o}F$) 
- t_min - Air temperature minimum of the day ($^{o}F$) 
- t_ave - Air temperature average of the day ($^{o}F$) 

Respond to the following questions:

4. Use the equation above to create a new column for the density of snow (*in percent*) for each day, and `print` or `cat` a statement showing the average: snow density, snow depth, swe, and average temperature for the time period. Make this a clean, organized statement with text indicating which values are being shown.

5. Create a single line plot showing both snow depth and density as a function of time (*use different colors for each and add the second plot using* `lines()`). Use the plot and equations above to describe the relationship between snow density and depth. 

6. Test if there is a relationship between snow water equivalent and precipitation accumulation and describe any relationship you observe. *simple visualizations and comparisons will work to explore relationships for now*

7. Based on your swe values, programatically determine the day in which new precipitation fell (i.e. the date in which a snow storm rolled in). Next, use the *swe* and *swe_med* (median historic swe value) to print the pre- and post-storm onset average percent median swe of the snowpack (i.e. on average how close, as a percentage, was swe before vs. after the storm started).

*Tip: the* `diff` *function is useful for exploring changes between neighboring values (like swe). Also, logic (e.g., ==, >, <=) can work with properly formatted dates to subset data.*

8. Consider the daily differences (`diff`) in *swe*, *depth*, and *density* for the snowpack over the ~30-day period. Use a well-formatted `print` or `cat` statement describing total increases and decreases in each of these variables. For example, some of these values may only increase over time, while others can increase and decrease (consider why!). Finally, to better understand the relationship between these variables, create one or more scatterplots to assess the relationships between the daily differences in: swe vs. depth, swe vs. density, and depth vs. density. (*Optional, statistically assess the relationship of each*)


<br />

##### **Extended snow and meteorological records**

Download and explore additional snow data using a new package.

Install the `snotelr` package with the `install.packages("snotelr")` command (you only need to install a package once), then run the following code to download all daily snotel data for the Timpanogos Divide site (same as above):

```r
# install.packages("snotelr")
library(snotelr)

# download data for a specific site (Timp Divide = 820)
snotel_download(site_id = 820,
                path = ".",      # can replace this with a specific file path
                internal = F)

# read in the dataset
site_820 <- read.csv("./snotel_820.csv") # modify filepath as needed
head(site_820)
```

9. Instead of the normal calendar year, hydrologists use something called a *water year* which typically begins on October 1st and ends September 30th. This is helpful, particularly in the western US, because it aligns with our winter snow season. Create a single line plot comparing last water year's *snow_water_equivalent* (in `"lightblue"`) to this season's (in `"black"`) over the water year day (e.g., if you are taking the class in Spring 2026, this seasons water year began Oct 1st, 2025).

*Tip: This will require you to subset the data by water year based on date ranges (e.g.,* `[date >= start_date & date <= end_date]`*), however, because these are different years, you cannot use the date as the common x-axis for the plot. One simple way to accomplish this would be to convert the dates in into a new variable (column) that indicates a numeric value for the "water year day of year," essentially just subtracting the dates from the first day (Oct 1, YEAR) of the water year to create a new variable.*

10. Switching gears, subset the data to only include values after Jan 1, 2010. Ignoring the water year, calculate and visualize (scatterplot) both the 1) annual maximum of *temperature_min* by year and 2) the annual maximum of *temperature_max* by year. Describe your observations and contextualize your results in relation to the snowpack (*Optional: quantitatively assess the strength of any relationships observed*)

*Tip: You accomplish this task in various ways, but one efficient way is to use* `aggregate` *or* `tapply` *to summarize values by year*


<br />

#### Submitting

Upload a single, organized **html file** written in **R Markdown** showing **questions, code solutions, and responses**. Be sure to support your responses with code and result outputs. Do not print or include unnecessary details in your final html document, *it should be easy to read and follow.*

*Please reach out if you have any questions or concerns!*

<br />

