---
title: "Module 7"
author: "Your Name"
output: html_document
---

```{r setup, include=FALSE} 
knitr::opts_chunk$set(echo = TRUE)
```

Provide complete code and written responses for each question below.

<br />

```{r echo=FALSE, warnings=F, message=F}
# read in packages
library(terra)
```

<br />

##### **Creating Spatial Data**

1. Using only the `terra` package and `vect()` function, transform this dataframe into a spatial vector data object and create a basic plot (`plot` function) showing the SNOTEL stations colored by elevation.

```{r}

```

Add additional code blocks as needed and provide any requested information...


<br />

2. Clean the station names to perform a successful join between the two datasets. Provide code to show your steps to clean and join the data and show that there are no `NA` values in your final joined dataset. 

```{r}

```


<br />

3. With your new joined dataset, create another basic spatial plot (with `terra` and `plot`) showing max SWE as `type = continuous`. Add the outline of the state of Utah using the `maps` package.

```{r}

```


<br />

4. (*1.5 pts*) Using the `terra` plot functions, create three plots side-by-side showing the max SWE at SNOTEL sites in Utah during the 2021-22 season. For methods, use 1) equal interval with 7 classes, 2) quantile, and 3) standard deviation. Specify the classification method the title of the legend in each plot and use sequential color schemes (one color of varying brightness) for equal interval and quantile, and a diverging color scheme (two colors) for the standard deviation classification method (see link above for more). Note the standard deviation scheme should include 1-3 standard deviations from the mean ($\mu \pm 3\sigma$). Provide 2-3 sentences describing the differences, benefits, and meaning of the classification methods used.

```{r}

```


<br />


#### **Manipulating Vector Data**

5. Create a basic map plot in `terra` that shows the individual watersheds colored by their unique names. Rather than adding a legend, use the `text()` function with the spatial data to add the names of each watershed to the map (*it is okay for this map to look a little messy*).

```{r}

```


<br />

6. Create an extent object (rectangular box) using the `ext()` function in terra for Utah Valley using the following: Latitude range: 39.915063 to 40.471646; Longitude range: -111.952211 to -111.536395. Plot the geometry (no variable) of the UT HUC layer and add the rectangular bounding box to the map in red, showing this as an area of interest. Add a legend in the top right of the map indicating what the area represents.

```{r}

```


<br />

7. Use the same workflow you used in the previous question along with the `crop()` function to clip the UT HUC watersheds data to only show the part of the state above 39 degrees Latitude. Create a map plot

```{r}

```


<br />

8. Create a subset of the UT_HUC vector that only contains the Great Salt Lake watershed. Then use the island vector above to cut the island out of the GSL watershed. Plot the new GSL feature that has a hole for Antelope Island, make the feature a color that would be used to represent water on a map. Add a text label to this plot indicating the location of Antelope Island.

```{r}

```


<br />

9. (*1.5 pts*) Create a new variables for the HUC4 and HUC2 codes, based on the HUC8 code. Dissolve the spatial vector into HUC4 units, then again in HUC2 units. Finally, create a map plot showing the three different levels of hydrologic units (HUC 8, 4, and 2), Use color, linewidth, linetype, and other arguments to distinguish between the different watershed units in the map plot.

```{r}

```


<br />
<br />

