---
title: "Module 8"
author: "Your Name"
output: html_document
---

```{r setup, include=FALSE} 
knitr::opts_chunk$set(echo = TRUE)
```

Provide complete code and written responses for each question below. **Use assignment questions for full details and tips.**

*Provide use figure and legend titles when appropriate*

<br />

```{r}
# read in packages
library(terra)
```

<br />


#### **Zonal relationships**

1. Use the `zonal` function to calculate average SWE (*values*) listed for the 2022 water year in each HUC6 in the state and plot this as interval (default) with 5 classes, using a color scale that varies in shade of blue (*e.g. blues9*). Run the `zonal` function again (or another tool) to calculate the total number of stations in each HUC6 and add this value with the `text` function in a complementary color (*e.g. red*). 

```{r}

```



2. Use the `expanse` function to calculate the area of each HUC6 unit in 10,000s of square kilometers (i.e. use `unit="km"` and multiply the area by `10e3` after). Use this value, along with your zonal statistics output of station counts by HUC6, to create a map that shows the density of stations over the area of each HUC6. How does this change your interpretation of SNOTEL representation within each HUC6?

```{r}

```

  
  
3. Convert your UT dem from meters of elevation to feet with simple arithmetic (*1 meter = 3.281 feet*) and use the elevation range for all SNOTEL sites in the dataset above to create a new dem (*dem_snotel*) that only shows cell values that are within the same range as SNOTEL sites state-wide. This can be accomplished with indexing and boolean logic, or the `clamp` function. Plot your subset elevation raster with the HUC6 borders overlapping. 

```{r}

```



4. Use both the UT DEM and your masked `dem_snotel` rasters to calculate the percentage (rounded to the nearest whole number) of pixels within the range of SNOTEL elevations by HUC6. To do this, use the `!is.na` function on the raster to create a binary raster. Then, use the `zonal` function to sum the number of pixels by HUC6 in each raster. Plot the percentages for each HUC6 with the default 5 interval classification (use a sequential color palette like `heat.colors` or similar to emphasize higher values), then add `text` values over the top that show the percentage followed by the "%" for each HUC6.

```{r}

```



5. Use the `extract` function to extract the raster UT dem elevation values (`x`) at each snotel site (`y`) in your vector point data (*note, the dimensions and order should match your snotel data, which can be added as a new column*). Use these values to create a scatterplot that shows predicted values (x-axis) with observed values. Add a 1:1 line using the `abline` function (*a perfect match would show the all points on the 1:1 line*). Add the RMSE (rounded to whole number) as text in the topleft corner of the plot.

```{r}

```


<br />

6. Using the code above and code used in question 4, recalculate the percent of pixels within station elevation ranges by HUC6. Create a side-by-side plot showing your map in question 4 next to this new modified version using the masked_dem calculated with the code above. Use the same color and range of values (breaks) between plots to maintain consistency. How does this change your interpretation of results?

```{r}

```


<br />

#### **Relationships between snow and elevation**


7. (*3 pts*) Investigate the affect of elevation on max seasonal SWE (*inches*) by HUC6 code.  Provide a two-by-three (six plots) window scatterplot showing the linear relationship for each HUC6 code with the equation and \R^2\ value listed for each (similar to the plot above). Secondly, list the significance (*p-value*) and the wellness of fit (\R^2\). How does the local linear relationship compare to the global relationship computed above?


```{r}

```


<br />
<br />

