---
title: "m8-extra-questions"
output: html_document
---





### Extra ideas for module

* elevation range within each HUC6 (rather than range for full state)
* find highest peak by HUC 6 and calculate the average distance to the peak for each station
* find distance between stations within each
* read in PRISM data for a final analysis


#### **highest peak (elevation) by HUC 6 and distance to station**

```{r}
# extract
vals <- extract(dut_ft, h6, cells = TRUE)

# Remove NA values
vals <- vals[!is.na(vals[,2]), ]

# Split by polygon ID and get max cell
max_cells <- do.call(rbind, 
                     lapply(split(vals, vals$ID), 
                            function(df) {
                              df[which.max(df[,2]), ]
}))

# Get xy coordinates of max cells
xy <- xyFromCell(dut_ft, max_cells$cell)

# Create SpatVector points
max_points <- vect(xy, crs = crs(dut_ft))

# Attach attributes (optional)
max_points$elev <- max_cells[,2]
max_points$poly_id <- max_cells$ID

plot(dut_ft)
plot(max_points, add=T, col='red')
plot(project(h6, dut_ft), add=T, border='grey85', lty=3)
plot(project(snotel_points, dut_ft), pch=3, add=T)
```


Calculate distance of all points

```{r}
distance(snotel_points, project(max_points, snotel_points))

nn <- nearby(snotel_points, project(max_points, snotel_points), k = 1)
head(nn)
dist = nn[,2]
sntl <- cbind(snotel_points, dist)
head(sntl)

# y variable (show average distance)
sno_peak_dist <- zonal(sntl, h6, mean, as.polygons=T, na.rm=T)
plot(sno_peak_dist, "y")
```


<br />
<br />

**EXTRA**

*Not fully formed - ignore for now*

Finally, use the Estimate (for *elevation_ft*) to describe how SWE changes relative to an increase of 10,000 feet within each of the six HUC6 regions.
    * *The estimate (also known as the regression coefficient) represents the expected change in the outcome variable (dependent variable) for a one-unit change in the predictor variable (independent variable), while holding all other predictors constant. In other words, this value will descibes how a 1,000 unit change in x (elevation in ft) impacts a 1 unit change in y (SWE in inches)*
    
    
```{r}
# extract statistic
swe_elev_est <- snotel_points_huc %>% as.data.frame() %>% 
  select(station, values, elevation_ft, HUC6) %>% 
  filter(HUC6 %in% keep_hucs$HUC6) %>% 
  mutate(values = values/12) %>% 
  group_by(HUC6) %>% 
  group_modify(
  ~ broom::tidy(lm(values ~ elevation_ft, data = .))
  ) %>% 
  filter(term=="elevation_ft") %>% 
  mutate(est_1000 = round(estimate*10000, 2)) %>%
  select(HUC6, est_1000)

# merge with hic 6 data
huc6_est <- merge(h6, swe_elev_est, by="HUC6")

# plot
plot(huc6_est, "est_1000", type='class', 
     main = "SWE increase (ft) per 10,000 ft gain in elevation",
     plg=list(title="estimate"))
text(huc6_est, "HUC6", col="white", face="bold", cex=0.7)
```